EPE and Speed Adaptive Extended Kalman Filter for Vehicle Position and Attitude Estimation with Low Cost GNSS and IMU Sensors

### 2014-04-29 Paper/main.tex
Vorstellung ITVS
Feedback ITVS

%\author{
%\IEEEauthorblockN{Paul Balzer}
%\IEEEauthorblockA{University of Applied Sciences\\
%Dresden, Germany\\
%balzer@htw-dresden.de}
%\and
%\IEEEauthorblockN{Toralf Trautmann}
%\IEEEauthorblockA{University of Applied Sciences\\
%Dresden, Germany\\
%trautmann@mw.htw-dresden.de}
%\and
%\IEEEauthorblockN{Oliver Michler}
%\IEEEauthorblockA{Technical University\\
%Dresden, Germany\\
%oliver.michler@tu-dresden.de}
%}

\author{
\IEEEauthorblockN{t.b.d.}
\IEEEauthorblockA{t.b.d.\\
t.b.d.}
\and
\IEEEauthorblockN{t.b.d.}
\IEEEauthorblockA{t.b.d.\\
t.b.d.}
\and
\IEEEauthorblockN{t.b.d.}
\IEEEauthorblockA{t.b.d.\\
t.b.d.}
}

@@

The implemented EKF estimates following state vector $x_k$, which is well known as the constant turn rate and velocity (CTRV) vehicle model + roll and pitch estimation:

\begin{equation}x_k= \left[ \begin{matrix}x \\ y \\ \psi \\ v \\ \dot \psi \\ \phi \\ \dot \phi \\ \Theta \\ \dot \Theta \end{matrix} \right] = \left[ \begin{matrix} \text{Position X (GNSS)} \\ \text{Position Y (GNSS)} \\ \text{Heading (GNSS)} \\ \text{Speed (GNSS)} \\ \text{Yaw Rate (IMU)} \\ \text{Pitch (IMU)} \\ \text{Pitchrate (IMU)} \\ \text{Roll (IMU)} \\ \text{Rollrate (IMU)} \end{matrix} \right]\end{equation}

@@

This paper lives this recommendation up and introduces an adaptively chosen weighting of roll and pitch angle and roll-/pitchrate, depending on the accelerations in x- or y-direction.

@@

The rollrate $\dot \Theta$, the pitchrate $\dot \phi$, the yawrate $\dot \psi$ as well as the speed $v$ are state control variables.

\begin{equation}\label{controlinput}u_k=\left[\begin{matrix}v & \dot\psi & \dot\phi & \dot\Theta\end{matrix}\right]^T\end{equation}

The vector $u$ drives the state from one timestep $k$ to the next one.

@@

The state transition function $g(x_k, u_k)$ is defined with
\begin{equation}\label{statetransitionfunction}
x_{k+1}=\left[\begin{matrix}x + \frac{v}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\\y + \frac{v}{\dot\psi} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)\\T \dot\psi + \psi\\v\\\dot\psi\\T \dot\phi + \phi\\\dot\phi\\T \dot\Theta + \Theta\\\dot\Theta\end{matrix}\right]
\end{equation}
for $\dot \psi \neq 0$. For straight driving without curvature, the state transition function is
\begin{equation}
x_{k+1}=\left[\begin{matrix}x + v T \cdot \cos(\psi)\\y + v T \cdot \sin(\psi)\\ \psi \\v\\\ 0 \\T \dot\phi + \phi\\\dot\phi\\T \dot\Theta + \Theta\\\dot\Theta\end{matrix}\right]
\end{equation}

@@

\begin{equation}Q=diag\label{Q}\left[\begin{matrix}\sigma_{v}^2 & \sigma_{{\dot\psi}}^2 & \sigma_{{\dot\phi}}^2 & \sigma_{{\dot\Theta}}^2 \end{matrix}\right]\end{equation}

@@

Assumptions for process noises for yawrate and velocity for a vehicle model are suggested in \cite{Kelly1994}. The process noise is best described with the question, how much the state can be propagated in one timestep by expected movement of the vehicle. So the maximal acceleration expected for a car might be $\SI{5}{\metre\per\square\second}$ under normal circumstances, so the $\sigma_v \approx a_\text{max}\cdot T$.
A typical maximal angular acceleration around vehicle z-axis might be $\SI{55}{\degree\per\square\second}$, which leads to a process noise of $\sigma_{\dot\psi} \approx \SI{55}{\degree\per\square\second} \cdot \frac{\pi}{180{,}0} \cdot T$.
The rotation around the pitch axis is much more dynamical, with typical angular accelerations of $\SI{200}{\degree\per\square\second}$. The rotation around the roll axis is in between and is assumed to be $\SI{130}{\degree\per\square\second}$.
The process noises for a $\SI{50}{\hertz}$ filter are listed in Table \ref{processnoisetable}.

\begin{table}[ht]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Process Noise Standard Deviations}
\label{processnoisetable}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{c l c}
\hline
Parameter & Decribtion & Value\\
\hline
$\sigma_v$ & Velocity Process Noise & \SI{0.1}{\metre\per\second} \\
$\sigma_{\dot \psi}$ & Yawrate Process Noise & \SI{0.019}{\radian\per\second}\\
$\sigma_{\dot \phi}$ & Pitchrate Process Noise & \SI{0.07}{\radian\per\second}\\
$\sigma_{\dot \Theta}$ & Rollrate Process Noise & \SI{0.045}{\radian\per\second}\\
\hline
\end{tabular}
\end{table}

@@

Because the estimation of roll \eqref{rollangle} and pitch \eqref{pitchangle} is only valid for quasistatic situations (which is acutally not valid for a moving vehicle), the standard deviation for the calculated attitude angles $\sigma_r$ are adaptive to the accelerations $a_x$ and $a_y$.

The novel approach, presented in this paper, is the very easy to implement and fast to calculate weighting of the measurement noise covariance matrix $R$.

\begin{equation}\label{R}R=diag\left[\begin{matrix}\sigma_{x}^2 & \sigma_{y}^2 & \sigma_{\phi}^2 & \sigma_{\Theta}^2 \end{matrix}\right]\end{equation}

\subsubsection{Position Measurement Uncertainty}

In every EKF filterstep, the standard deviations for $\sigma_x$ and $\sigma_y$ are calculated, depending on the speed and additionally depending on the estimated position error ($EPE$), which is provided by the GNSS modul itself.
\begin{equation}\sigma_x^2 = \sigma_y^2 = \sigma_\text{v}^2 + \sigma_\text{EPE}^2\end{equation}
with
\begin{equation}\label{sigmav}\sigma_v = (v+\epsilon)^{-\xi}\end{equation}
\begin{equation}\label{sigmaepe}\sigma_\text{EPE} = \zeta \cdot EPE\end{equation}

The variables $\epsilon$, $\xi$ and $\zeta$ are tuneable parameters.

@@

Because of the simplifications \eqref{deltax} and \eqref{deltay}, the Jacobian of the measurement function $h$ with respect to the state $x_k$ is simple and computationally fast with 
\begin{equation}\label{hgps}J_H=\left[\begin{matrix}1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0\end{matrix}\right]\end{equation}
when a new GNSS position measurement is available. In the practical implementation, the GNSS provide position information with \SI{10.0}{\hertz} and the EKF is estimating with \SI{50.0}{\hertz}. If no GNSS position information is available, the Jacobian of the measurement function $h$ with respect to the state $x_k$ is
\begin{equation}J_H=\left[\begin{matrix}0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0\end{matrix}\right]\end{equation}

@@

To evaluate the adaptive EKF, a typical urban scenario, with shading from a buildings, as well as a vehicle stop, was simulated (see Fig. \ref{Testdata}).

@@

The car starts at $x=0$, $y=0$ with $v=\SI{50.0}{\kilo\metre\per\hour}$ and decelerates until it stops. It is standing for $\SI{20}{\second}$ and accelerating until it reaches $v=\SI{50.0}{\kilo\metre\per\hour}$ again. Then it is passing a building, which disrupts the signal quality of the GNSS, so the position measurement is changed by $+\SI{15}{\metre}$ in $y$-direction, the $EPE$ is raised as well.

@@

The summand $\epsilon$ for \eqref{sigmav} is initialized with $\SI{1.0}{\metre\per\second}$, the exponent $\xi$ for \eqref{sigmav} with $500{,}0$ and the factor $\zeta$ for \eqref{sigmaepe} with $50{,}0$. The resulting values for $\sigma_x^2$ and $\sigma_y^2$ for a range of velocities and $EPE$s are shown in Fig. \ref{R1}.

@@

These parameters are empirically chosen and are subject to change for different cars or other driving scenarios or street quality.

@@

As one can see in Fig. \ref{CTRV-Position-Testdata}, the filter follows the trajectory of the GNSS measurements. If the EPE raises, the filter is more willing to trust the control input (see \eqref{controlinput}) instead of the position measurements.

@@

The sum of the square of the CTE for the simulation is a value to quantify the filter performance with respect to the correct trajectory estimation (see Fig. \ref{CTRV-CTE-Testdata}).

\begin{figure}[ht]
\centering
\includegraphics[width=3.5in]{images/Extended-Kalman-Filter-CTRV-CTE-Testdata}
\caption{Cross-Track-Error (not position estimate) with sum of the square}
\label{CTRV-CTE-Testdata}
\end{figure}

@@

The standard EKF was set up with static $\sigma_x^2=\SI{36}{\square\metre}$ and $\sigma_y^2=\SI{36}{\square\metre}$ values. The result is shown in Fig.  \ref{Boxplot}.

@@

In comparison, the adaptive EKF reduces the $\Sigma \text{CTE}_x^2$ by $\SI{47}{\percent}$, the $\Sigma \text{CTE}_y^2$ by $\SI{59}{\percent}$, which for sure depends on the driving direction as well as on the chosen parameters.

@@

To test the performance of the filter with real GNSS data as well as a real car, the following sensores were used: LSM303   3-axis accelerometer and 3-axis magnetometer, ITG-3200 3-axis gyro, PA6H GNSS receiver. The sensors are available with the Tinkerforge IMU Brick\footnote{\url{http://www.tinkerforge.com/en/doc/Hardware/Bricks/IMU_Brick.html}} and GPS Bricklet\footnote{\url{http://www.tinkerforge.com/en/doc/Hardware/Bricklets/GPS.html}}.

For vehicle speed lower than $\SI{1,0}{\kilo\metre\per\hour}$ (GNSS), the speed and yawrate are set to zero, because the Doppler effect doesn't work for speed estimation in the GNSS receiver and turning while standing still is very unlikely.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=3.4in]{images/Extended-Kalman-Filter-CTRV-Position}
\caption{Performed test drive with adaptive values of measurement uncertainty values for matrix $R$ (color)}
\label{ctrv-position}
\end{figure}

@@

As one can see, the stop before the last left turn got a high position measurement uncertainty, which in detail is shown in figure \ref{ctrv-position-detail}.

\begin{figure}[ht]
\centering
\includegraphics[width=3.2in]{images/Extended-Kalman-Filter-CTRV-Position-Detail}
\caption{Stop within the performed test drive with raised uncertainty for $R$ and position estimation from adaptive EKF}
\label{ctrv-position-detail}
\end{figure}

The state variables for vehicle speed and yawrate are nearly perfectly aligned, because they fit very well and they are state control variables as well. The state is forced to follow them.

While standing still, the heading measurement of the GNSS receiver is inaccurate (see Fig. \ref{ctrv-states} between filter step $k\approx2100\dotsc2300$). The filter performes very well on this situation and keeps the heading in the correct direction. The EKF position estimation (see Fig. \ref{ctrv-position-detail}) is not disturbed by inaccurate GNSS measurements while standing still and is dynamically responsive while driving.

For the conducted test drive, the values for the measurement uncertainties in $R$ are calculated like shown in figure \ref{adaptive-R}.

@@

\section{Conclusion}
In this paper we presented a novel approach to adaptively calculate the measurement uncertainty for an improved position and attitude estimation, based on the Estimated Position Error and the speed, determined by the low cost GNSS receiver.
We explained the basics and used some empirically chosen values to parametrize the filter. With that, we showed with simulated data, that the filter performes significantly better than a standard Extended Kalman Filter. Based on that, we conducted test drives and used the developed Adaptive EKF for a real world dataset, which proved its ability to improve the position estimation with partly bad signal quality. In addition to that, the filter also performs pretty well on dynamic situations and is not loosing the ability to follow dynamic vehicle movements.
The filter cannot be used to get rid of bias in position estimation, because the EPE from GNSS has, by definition, no information about static drift of the position information.
The presented filter can be used to get significantly better results while standing still or driving slowly as well as keeping the heading fixed while do so.
Additionally, the attitude of the vehicle is estimated, based on the filter presented in \cite{Madgwick2010} in rest position and with angular velocity sensors from IMU while driving.

@@

%The authors would like to thank the Free State of Saxony and the European Union, which funded the research from the ESF fond.
T.b.d.

\appendices
\section{Programm Code and EKF Implementation}
The implemenented EKF code as well as all figures and the data used in this paper can be found online at t.b.d %\url{http://balzer82.github.com/MMAR14}

@@

\begin{equation}
J_A=\left[\begin{matrix}1 & 0 & J_{A,13} & J_{A,14} & J_{A,15} & 0 & 0 & 0 & 0\\0 & 1 & J_{A,23} & J_{A,24} & J_{A,25} & 0 & 0 & 0 & 0\\0 & 0 & 1 & 0 & T & 0 & 0 & 0 & 0\\0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 1 & T & 0 & 0\\0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0\\0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & T\\0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\end{matrix}\right]
\end{equation}

@@

\begin{equation}J_{A,13}=\frac{v}{\dot\psi} \left(- \cos{\left (\psi \right )} + \cos{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,14}=\frac{1}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,15}=\frac{T v}{\dot\psi} \cos{\left (T \dot\psi + \psi \right )} - \frac{v}{\dot\psi} \cdot J_{A,14}\end{equation}
\begin{equation}J_{A,23}=\frac{v}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,24}=\frac{1}{\dot\psi} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,25}=\frac{T v}{\dot\psi} \sin{\left (T \dot\psi + \psi \right )} - \frac{v}{\dot\psi}\cdot J_{A,24}\end{equation}

@@

\begin{equation}J_G=\left[\begin{matrix}J_{G,11} & J_{G,12} & 0 & 0\\ J_{G,21} & J_{G,22} & 0 & 0\\0 & T & 0 & 0\\1 & 0 & 0 & 0\\0 & 1 & 0 & 0\\0 & 0 & T & 0\\0 & 0 & 1 & 0\\0 & 0 & 0 & T\\0 & 0 & 0 & 1\end{matrix}\right]\end{equation}
with
\begin{equation}J_{G,11}=\frac{1}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{G,12}=\frac{T v}{\dot\psi} \cos{\left (T \dot\psi + \psi \right )} - \frac{v}{\dot\psi} \cdot J_{G,11}\end{equation}
\begin{equation}J_{G,21}=\frac{1}{\dot\psi} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{G,22}=\frac{T v}{\dot\psi} \sin{\left (T \dot\psi + \psi \right )} - \frac{v}{\dot\psi} \cdot J_{G,21}\end{equation}

>>>>

\author{
\IEEEauthorblockN{Paul Balzer}
\IEEEauthorblockA{University of Applied Sciences\\
Dresden, Germany\\
balzer@htw-dresden.de}
\and
\IEEEauthorblockN{Toralf Trautmann}
\IEEEauthorblockA{University of Applied Sciences\\
Dresden, Germany\\
trautmann@mw.htw-dresden.de}
\and
\IEEEauthorblockN{Oliver Michler}
\IEEEauthorblockA{Technical University\\
Dresden, Germany\\
oliver.michler@tu-dresden.de}
}

%\author{
%\IEEEauthorblockN{t.b.d.}
%\IEEEauthorblockA{t.b.d.\\
%t.b.d.}
%\and
%\IEEEauthorblockN{t.b.d.}
%\IEEEauthorblockA{t.b.d.\\
%t.b.d.}
%\and
%\IEEEauthorblockN{t.b.d.}
%\IEEEauthorblockA{t.b.d.\\
%t.b.d.}
%}

@@

The implemented EKF estimates following state vector $x_k$, which is well known as the constant turn rate and velocity (CTRV) vehicle model plus additionally roll and pitch estimation:

\begin{equation}\label{state}x_k= \left[ \begin{matrix} x\\y\\ v \\ \psi\\\phi\\\Theta \end{matrix}\right] = \left[ \begin{matrix} \text{Position x (GNSS)}\\ \text{Position y (GNSS)}\\ \text{Speed (GNSS)} \\ \text{Heading (GNSS)} \\ \text{Pitch (IMU)} \\ \text{Roll (IMU)} \end{matrix}\right]\end{equation}

@@

This paper lives this recommendation up and introduces an adaptively chosen weighting of roll and pitch angle, depending on the accelerations in lateral or longitudinal direction.

@@

The longitudinal acceleration $a_x$, rollrate $\dot \Theta$, pitchrate $\dot \phi$ and the yawrate $\dot \psi$ are state control variables.

\begin{equation}\label{controlinput}u_k=\left[\begin{matrix}a_x & \dot\psi & \dot\phi & \dot\Theta \end{matrix}\right]^T\end{equation}

They are acquired by the IMU. The proposed filter could be used without any physical connection to the vehicle sensors (e.g. CAN-Bus) and therefore it is useful for mobile measurement systems. The state control vector $u$ consists only of values, acquired by the IMU, so the filter can run with high update rate in dead reckoning mode without GNSS information and the measurements from GNSS are used with lower update rate as correction.

@@

The state transition function $g(x_k, u_k)$ is defined with
\begin{equation}\label{statetransitionfunction}
x_{k+1}=\left[\begin{matrix}x + \frac{v}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\\y + \frac{v}{\dot\psi} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)\\T a_{x} + v\\T \dot\psi + \psi\\T \dot\phi + \phi\\T \dot\Theta + \Theta\end{matrix}\right]
\end{equation}

for $\dot \psi \neq 0$. For straight driving without curvature, the state transition function is

\begin{equation}
x_{k+1}=\left[\begin{matrix}x + vT \left(\cos(\psi)\right)\\y + vT \left(\sin (\psi) \right) \\ T a_{x} + v\\\psi \\T \dot\phi + \phi\\T \dot\Theta + \Theta\end{matrix}\right]
\end{equation}

@@

\begin{equation}Q=diag\label{Q}\left[\begin{matrix}\sigma_{a}^2 & \sigma_{{\dot\psi}}^2 & \sigma_{{\dot\phi}}^2 & \sigma_{{\dot\Theta}}^2 \end{matrix}\right]\end{equation}

@@

Assumptions for process noises for a vehicle model are suggested in \cite{Kelly1994}. The process noise is best described with the question, how much the state can be propagated in one timestep by expected movement of the vehicle. So the jerk expected for a car might be $\SI{300}{\metre\per\cubic\second}$ under normal circumstances, so the $\sigma_a \approx \dot a_\text{max}\cdot T$.

A typical maximal angular acceleration around vehicle z-axis might be $\SI{80}{\degree\per\square\second}$, which leads to a process noise of $\sigma_{\dot\psi} \approx \SI{80}{\degree\per\square\second} \cdot \frac{\pi}{180{,}0} \cdot T$.
The rotation around the pitch axis is much more dynamically, with typical angular accelerations of $\SI{200}{\degree\per\square\second}$ because of bumps and road surface quality. The angular acceleration around the roll axis is assumed to be $\SI{200}{\degree\per\square\second}$, too.

The process noises for a $\SI{50}{\hertz}$ filter are listed in Table \ref{processnoisetable}.

\begin{table}[ht]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Process Noise Standard Deviations}
\label{processnoisetable}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{c l c}
\hline
Parameter & Decribtion & Value\\
\hline
$\sigma_a$ & Acceleration Process Noise & \SI{6.0}{\metre\per\square\second} \\
$\sigma_{\dot \psi}$ & Yawrate Process Noise & \SI{0.028}{\radian\per\second}\\
$\sigma_{\dot \phi}$ & Pitchrate Process Noise & \SI{0.070}{\radian\per\second}\\
$\sigma_{\dot \Theta}$ & Rollrate Process Noise & \SI{0.070}{\radian\per\second}\\
\hline
\end{tabular}
\end{table}

@@

The attitude estimation of the IMU filter \cite{Madgwick2010}, which is used as measurement for roll \eqref{rollangle} and pitch \eqref{pitchangle}, is only valid for quasistatic situations (which is acutally not valid for a moving vehicle), the standard deviation for the calculated attitude angles are adaptive to the accelerations $a_x$ for pitch and $a_y$ for roll.

The novel approach, presented in this paper, is the very easy to implement and fast to calculate elements of the measurement noise covariance matrix $R$.

\begin{equation}\label{R}R=diag\left[\begin{matrix}\sigma_{x}^2 & \sigma_{y}^2 & \sigma_{v}^2 & \sigma_{\psi}^2 & \sigma_{\phi}^2 & \sigma_{\Theta}^2 \end{matrix}\right]\end{equation}

\subsubsection{Position Measurement Uncertainties}

In every EKF filterstep, the standard deviations for $\sigma_x$ and $\sigma_y$ are calculated, depending on the speed and additionally depending on the estimated position error ($EPE$), which is provided by the GNSS modul itself.
\begin{equation}\sigma_x^2 = \sigma_y^2 = c \cdot \sigma_\text{v}^2 + \sigma_\text{EPE}^2\end{equation}
with
\begin{equation}\label{sigmav}\sigma_v = (v+\epsilon)^{-\xi}\end{equation}
\begin{equation}\label{sigmaepe}\sigma_\text{EPE} = \zeta \cdot EPE\end{equation}

The variables $\epsilon$, $\xi$ and $\zeta$ are tuneable parameters. Factor $c=s^2$ for unit correction.

@@

Because of the simplifications \eqref{deltax} and \eqref{deltay}, the Jacobian of the measurement function $h$ with respect to the state $x_k$ is simple and computationally fast with 
\begin{equation}\label{hgps}J_H=diag\left[\begin{matrix}1 & 1 & 1 & 1 & 1 &1\end{matrix}\right]\end{equation}
when a new GNSS position measurement is available. In the practical implementation, the GNSS provide position information with \SI{10.0}{\hertz} and the EKF is estimating with \SI{50.0}{\hertz} (IMU measurement frequency). If no GNSS position information is available, the correspondig elements in $J_H$ are zero.

@@

To evaluate the adaptive EKF, a typical urban scenario, with shading from a building, as well as a vehicle stop, was simulated (see Fig. \ref{Testdata}).

@@

The car starts at $x=0$, $y=0$ with $v=\SI{50.0}{\kilo\metre\per\hour}$ and decelerates until it stops. It is standing for $\SI{20}{\second}$ and accelerating until it reaches $v=\SI{50.0}{\kilo\metre\per\hour}$ again. Then it is passing a building, which corrupts the signal quality of the GNSS, and the position measurement is disrupted by $+\SI{15}{\metre}$ in $y$-direction.

@@

The summand $\epsilon$ for \eqref{sigmav} is initialized with $\SI{0.1}{\metre\per\second}$, the exponent $\xi$ for \eqref{sigmav} with $500{,}0$ and the factor $\zeta$ for \eqref{sigmaepe} with $50{,}0$. The resulting values for $\sigma_x^2$ and $\sigma_y^2$ for a range of velocities and $EPE$s are shown in Fig. \ref{R1}.

@@

These parameters are empirically chosen and are subject to change for different cars or other driving scenarios or street qualities.

@@

As one can see in Fig. \ref{CTRV-Position-Testdata}, the filter follows the trajectory of the GNSS measurements. If the $EPE$ raises, the filter is more willing to trust the control input instead of the position measurements.

@@

The sum of the square of the CTE over the whole dataset is a value to quantify the filter performance with respect to the correct trajectory estimation.

@@

The standard EKF was set up with static $\sigma_x^2=\SI{144}{\square\metre}$, $\sigma_y^2=\SI{144}{\square\metre}$ values. The result is shown in Fig.  \ref{Boxplot}.

@@

In comparison, the adaptive EKF reduces the $\Sigma \text{CTE}_{x,y}^2$ by around $\SI{80}{\percent}$, which for sure depends on the driving direction as well as on the chosen parameters.

@@

\subsection{Low Cost Sensors}

The measurement and control data of the filter were logged by a LSM303 3-axis accelerometer and 3-axis magnetometer, ITG-3200 3-axis gyro and PA6H GNSS receiver. The sensors are available with the Tinkerforge IMU Brick\footnote{\url{http://www.tinkerforge.com/en/doc/Hardware/Bricks/IMU_Brick.html}} and GPS Bricklet\footnote{\url{http://www.tinkerforge.com/en/doc/Hardware/Bricklets/GPS.html}}.

All sensors are mobile, so no connection to the car is neccessary, but could improve the estimation.

\subsection{Ground Truth GNSS Sensor}

The ground truth was logged with a multifrequency aerial antenna (JAVAD GrAnt-G3T) and -receiver (JAVAD Delta), corrected with a virtual reference station (VRS) and Satellite Positioning Service of the German State Survey (SAPOS) data via GPRS modem (come2ascos GenPro).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=3.4in]{images/Extended-Kalman-Filter-CTRV-Position}
\caption{Performed test drive with adaptive values of measurement uncertainty values for matrix $R$. (GNSS measurement dots are reduced to every 10th for better visualisation in this figure)}
\label{ctrv-position}
\end{figure}

@@

As one can see, the stop before the right turn got a high position measurement uncertainty, which in detail is shown in figure \ref{ctrv-position-detail}.

\begin{figure}[ht]
\centering
\includegraphics[width=3.2in]{images/Extended-Kalman-Filter-CTRV-Position-Detail}
\caption{Detailed view of the stop within the performed test drive with raised uncertainty for $R$. (GNSS measurement dots are reduced to every 10th for better visualisation in this figure)}
\label{ctrv-position-detail}
\end{figure}

While standing still, the heading measurement of the GNSS receiver is inaccurate (see Fig. \ref{ctrv-states} between filter step $k\approx2200\dotsc2400$). The filter performes very well on this situation and keeps the heading in the correct direction. The EKF position estimation (see Fig. \ref{ctrv-position-detail}) is not disturbed by inaccurate GNSS measurements while standing still and is dynamically responsive while cornering. The low cost GNSS equipment with the proposed adaptive Extended Kalman Filter delivers a robust position estimation, which alignes with the ground truth measurement.

The values for the measurement uncertainties in $R$ are calculated like shown in figure \ref{adaptive-R}.

@@

\section{Conclusion}
In this paper we presented a novel approach to adaptively calculate the measurement uncertainty for an improved position and attitude estimation, based on the Estimated Position Error and the speed, determined by the low cost GNSS receiver.
We explained the basics and used some empirically chosen values to parametrize the filter. With that, we showed with simulated data, that the filter performes significantly better than a standard Extended Kalman Filter. Based on that, we conducted test drives and used the developed Adaptive EKF for a real world dataset, which proved its ability to improve the position estimation with partly bad signal quality. In addition to that, the filter also performs pretty well on dynamic situations and is not loosing the ability to follow dynamic vehicle movements.
The filter cannot be used to get rid of bias in position estimation, because the EPE from GNSS has, by definition, no information about static drift of the position information.
The presented filter can be used to get significantly better results while standing still or driving slowly as well as keeping the heading fixed while do so.
Additionally, the attitude of the vehicle is estimated, based on the filter presented in \cite{Madgwick2010}.

@@

The authors would like to thank the Free State of Saxony and the European Union, which funded this research from the ESF fond.
%T.b.d.

\appendices
\section{Programm Code and EKF Implementation}
The implemenented EKF code as well as all figures and the data used in this paper can be found online at ... %\url{http://balzer82.github.com/MMAR14}

@@

\begin{equation}\left[\begin{matrix}1 & 0 & J_{A,13} & J_{A,14} & 0 & 0\\0 & 1 & J_{A,23} & J_{A,24} & 0 & 0\\0 & 0 & 1 & 0 & 0 & 0\\0 & 0 & 0 & 1 & 0 & 0\\0 & 0 & 0 & 0 & 1 & 0\\0 & 0 & 0 & 0 & 0 & 1\end{matrix}\right]\end{equation}

@@

\begin{equation}J_{A,13}=\frac{1}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,14}=\frac{v}{\dot\psi} \left(- \cos{\left (\psi \right )} + \cos{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,23}=\frac{1}{\dot\psi} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)\end{equation}
\begin{equation}J_{A,24}=\frac{v}{\dot\psi} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)\end{equation}

@@

\begin{equation}J_G=\left[\begin{matrix}0 & J_{G,12} & 0 & 0\\0 & J_{G,22} & 0 & 0\\T & 0 & 0 & 0\\0 & T & 0 & 0\\0 & 0 & T & 0\\0 & 0 & 0 & T\end{matrix}\right]\end{equation}

\begin{equation}
	\begin{split}
    	J_{G,12}=\frac{T v}{\dot\psi} \cos{\left (T \dot\psi + \psi \right )} - \\ 		\frac{v}{\dot\psi^{2}} \left(- \sin{\left (\psi \right )} + \sin{\left (T \dot\psi + \psi \right )}\right)
	\end{split}
\end{equation}

\begin{equation}
	\begin{split}
    	J_{G,22}=\frac{T v}{\dot\psi} \sin{\left (T \dot\psi + \psi \right )} - \\ \frac{v}{\dot\psi^{2}} \left(\cos{\left (\psi \right )} - \cos{\left (T \dot\psi + \psi \right )}\right)
    \end{split}
\end{equation}
